%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%

% Note to Jared

\section{Note to Jared}

This is my first attempt at composing a "Design Description Document", and it is different than a lot of documents (such as proposals) in biology.
As I understand it, the purpose of the document is to describe in detail \textbf{1) the goal of the project, 2) the specific requirements to reach that goal, and 3) the design decisions for specific features to meet those requirements}.
In this type of project there are a lot of decisions about details that matter and make a big difference (such as whether or not to host the tool on a server).
The idea is that making those decisions early on and getting feedback will help to make them better.
Also the thorough plan for the project will help the project succeed towards a specific goal.
That said, this document does not even come close to addressing all the details and decisions for this project.
It will be necessary to make a lot of those decisions as we get to them, since problems and opportunities will probably arise during implementation.


\vspace{1cm}


\textbf{Please consider the following subjects as you critique the document.}
\\ \textbf{1)} How valuable is the goal of the project? How well does the document communicate this goal?
\\ \textbf{2)} How accurately does the document describe the user? How well does the document communicate the requirements of the project with respect to the user? What requirements would you include or exclude?
\\ \textbf{3)} How appropriate are the design features to support the study of metabolism? What features would you include or exclude?

\vspace{1cm}

\textbf{Thank you} very much for reading the draft and giving feedback!

\vspace{1cm}

\textbf{To Do...}

\vspace{1cm}

\textbf{1) Figures}
\\ This draft does not yet have \textbf{figures to illustrate the design of the graphical interface}.
Planning out the graphical design of the interface will be necessary and important for the project, so it'll be worth it to spend the time now so that I can get feedback.
Designing figures will be my next step on this document, and it could be a lot of work.

\textbf{2) Schedule}
\\ This draft's description of the implementation steps and schedule in the \textbf{"Schedule"} section is incomplete.
This section is tricky, since it breaks the implementation process down into detailed steps.
It's also difficult to anticipate accurately how long each step will take.

% Page break
\newpage



